---
title: EDA 00.1: Sounder's Data Prep & Exploration
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data

## only working with the Sounders' data in this markdown file

```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_cleaned_sep.csv")

```

# Data cleaning

## NOTE: this process only needs to be completed again if you are starting with a new baseline data file. The steps below are the process to complete the tasks listed above each block - we write out the files to "\~/output/data_adjusted/filename.csv" so that we don't have to keep doing these steps.

## clean up the df, fill in the empty data points, fix variation in capitalization of comments, channel & source

```{r}

# before importing the data, I changed all column names (for columns we're going to use) to lowercase or camelcase because I hate remembering to capitalize when I don't have to.

# remove the columns we're not going to use
columns_to_drop <- c("UTMx", "UTMy", "ActLat", "ActLong", "species", "direction", "reportingParty")
data <- data %>% select(-all_of(columns_to_drop))

# take care of the empties with predetermined values
data$time[data$time == ""] <- "0:00"
data$count[data$count == ""] <- 1
data$count[is.na(data$count)] <- 1
data$channel[data$channel == ""] <- "unknown"
data$source[data$source == ""] <- "unknown"

# fix variable names in the channel and source columns. 
# since both uppercase and lowercase are used, they will be treated as unique cases, this step stops that from happening. The 'dplyr' package does this.
data <- data %>%
  mutate(channel = tolower(channel),
         source = tolower(source),
         comments = tolower(comments))

# check your work & then write out the new df as a safety stop
head(data)
#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_cleaned.csv", row.names = FALSE)

```

## format sightDate and pull out the month from the date

```{r}

# convert SightDate to date format and extract the month
data <- data %>%
  mutate(sightDate = as.Date(sightDate, format = "%m/%d/%y"),
         month = format(sightDate, "%m"))
         
```

## add a season column

```{r}

# create a season column
data <- data %>%
  mutate(season = case_when(
    month(sightDate) %in% c(12, 1, 2) ~ "winter",
    month(sightDate) %in% c(3, 4, 5) ~ "spring",
    month(sightDate) %in% c(6, 7, 8) ~ "summer",
    month(sightDate) %in% c(9, 10, 11) ~ "fall"
  ))

# double check your work and save it - safety stop; the file name is to help me keep track of what was saved at which step
head(data)
#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_clean_noIdentifiers.csv", row.names = FALSE)

```

## list identified whales from the comment column, make a new column


# EDA

## total count of Sounders sighted: 2742
```{r}

# total number of sounders sighted from 2013 - 2023
# save the table it creates if you don't want to copy paste the data OR you can knit this markdown file to a pdf or html file that supports printing/ manipulation.

total_count <- sum(data$count, na.rm = TRUE)
print(paste("Total Whale Count:", total_count))

```

## count by year
## 2018, 2019 and 2022 are the top years for Sounders' sightings with 445, 430 and 413 respectively
```{r}

yearly_counts <- data %>%
  group_by(year) %>%
  summarise(yearlyCount = sum(count, na.rm = TRUE))

# print
print(yearly_counts)

```

## count by month
## April and March are the most visited months, 1011 and 866 sightings respectively.
```{r}

# total number of whales counted per month regardless of year

monthly_counts <- data %>%
  group_by(month) %>%
  summarise(totalCount = sum(count, na.rm = TRUE))

print(monthly_counts)

```

## count by season
## Spring is the big visiting season by far with 2217 sightings.
```{r}

# group by season
seasonal_counts <- data %>%
  group_by(season) %>%
  summarise(totalCount = sum(count, na.rm = TRUE)) %>%
  arrange(match(season, c("Winter", "Spring", "Summer", "Fall")))

# print
print(seasonal_counts)

# plot seasonal trends
library(ggplot2)

ggplot(seasonal_counts, aes(x = season, y = totalCount, fill = season)) +
  geom_bar(stat = "identity") +
  labs(title = "Whale Sightings by Season",
       x = "Season",
       y = "Total Count of Sightings") +
  theme_minimal()

```

## count by season and year

```{r}

library(ggplot2)
library(dplyr)

# Group by year and season to get seasonal counts for each year
seasonal_counts_by_year <- data %>%
  group_by(year, season) %>%
  summarise(totalCount = sum(count, na.rm = TRUE)) %>%
  arrange(year, match(season, c("Winter", "Spring", "Summer", "Fall")))

# Print the grouped data
print(seasonal_counts_by_year)

# Plot seasonal trends by year
ggplot(seasonal_counts_by_year, aes(x = season, y = totalCount, fill = season)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ year) +  # Create a grid of plots, one for each year
  labs(title = "Whale Sightings by Season and Year",
       x = "Season",
       y = "Total Count of Sightings") +
  theme_minimal()

```

## count by quad
## Quad 384 is the big winner by several hundered sightings
```{r}

# total number of whales counted per quad regardless of year, descending order
quad_counts <- data %>%
  group_by(quad) %>%
  summarise(totalCount = sum(count, na.rm = TRUE))%>%
  arrange(desc(totalCount))

print(quad_counts)

```

## 'map' of sightings

```{r}

# plotting with lat/ long coordinates, but not overlayed on a map - just want to see what it looks like

library(ggplot2)

ggplot(data, aes(x = long, y = lat, color = whale_identifiers)) +
  geom_point(alpha = 0.6) +
  labs(title = "Map of Whale Sightings", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```

## timeline of sightings

```{r}

# convert sightDate to Date format (if it wasn't done earlier)
#timeline_data$sightDate <- as.Date(timeline_data$sightDate, format = "%m/%d/%y")

ggplot(timeline_data, aes(x = sightDate, y = whale_identifiers)) +
  geom_point(aes(color = whale_identifiers), alpha = 0.6) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Timeline of Whale Sightings", x = "Date", y = "Whale Identifiers", color = "Whale Identifiers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## movement patterns

```{r}

# Sort by identifiers and date
timeline_data <- timeline_data[order(timeline_data$whale_identifiers, timeline_data$sightDate), ]

ggplot(timeline_data, aes(x = long, y = lat, group = whale_identifiers, color = whale_identifiers)) +
  geom_path(alpha = 0.6) +
  geom_point(alpha = 0.6) +
  labs(title = "Movement Patterns of Identifiable Whales", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```
