---
title: Sounder's Data Prep
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")

```

# Data cleaning
## clean up the df, fill in the empty data points, fix variation in capitalization of comments, channel & source
```{r}

# before importing the data, I changed all column names (for columns we're going to use) to lowercase or camelcase because I hate remembering to capitalize when I don't have to.

# remove the columns we're not going to use
columns_to_drop <- c("UTMx", "UTMy", "ActLat", "ActLong", "species", "direction", "reportingParty")
data <- data %>% select(-all_of(columns_to_drop))

# take care of the empties with predetermined values
data$time[data$time == ""] <- "0:00"
data$count[data$count == ""] <- 1
data$count[is.na(data$count)] <- 1
data$channel[data$channel == ""] <- "unknown"
data$source[data$source == ""] <- "unknown"

# fix variable names in the channel and source columns. 
# since both uppercase and lowercase are used, they will be treated as unique cases, this step stops that from happening. The 'dplyr' package does this.
data <- data %>%
  mutate(channel = tolower(channel),
         source = tolower(source),
         comments = tolower(comments))

# check your work & then write out the new df as a safety stop
head(data)
#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/data_cleaned.csv", row.names = FALSE)

```

## format sightDate and pull out the month from the date
```{r}

# convert SightDate to date format and extract the month
data <- data %>%
  mutate(sightDate = as.Date(sightDate, format = "%m/%d/%y"),
         month = format(sightDate, "%m"))
         
```

## add a season column
```{r}

# create a season column
data <- data %>%
  mutate(season = case_when(
    month(sightDate) %in% c(12, 1, 2) ~ "Winter",
    month(sightDate) %in% c(3, 4, 5) ~ "Spring",
    month(sightDate) %in% c(6, 7, 8) ~ "Summer",
    month(sightDate) %in% c(9, 10, 11) ~ "Fall"
  ))

```

## list identified whales from the comment column, make a new column
```{r}

# we're going to extract all of the comments from the column and make them individual rows so we can assess what we have. This code came directly from ChatGPT

library(dplyr)
library(tidyr)
library(stringr)

# Function to extract whale identifiers, including plain numbers and case-insensitive CRC identifiers
# Updated pattern:
  # 1. Matches # followed by digits (#123)
  # 2. Matches case-insensitive CRC followed by digits (CRC-123, crc123)
  # 3. Matches plain numbers (1234)
  # 4. Matches names with one or two capitalized words (e.g., Shackleton, Little Patch)
extract_whale_identifiers <- function(comment) {
  pattern <- "(#[\\d]+|(?:[Cc][Rr][Cc][-]?\\d+)|\\b\\d+\\b|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"
  str_extract_all(comment, pattern)
}

# Apply the extraction function to create a new column
data <- data %>%
  mutate(whale_identifiers = sapply(lapply(comments, extract_whale_identifiers), paste, collapse = ", "))


# Filter for rows where whale_identifiers is not NA
#timeline_data <- data %>% filter(!is.na(whale_identifiers))

# check your work & write out as a table to save progress if you need to.
head(data)

#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/timeline_data.csv", row.names = FALSE)

```

## list identified whales, unique names
```{r}

# we have lists of IDs in the comments column, but not all of them are actual IDs but rather observation note (i.e., young whale) and we want to know what we're working with. This code came from ChatGPT.

library(stringr)

# Define the regex pattern used for extraction
pattern <- "(#[\\d]+|CRC-\\d+|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"

# Extract identifiers from comments
extracted_identifiers <- str_extract_all(data$comments, pattern)

# Flatten the list and get unique values
unique_identifiers <- unique(unlist(extracted_identifiers))

# View the unique identifiers
print(unique_identifiers)

# write out to your data_adjusted output folder

# unblock this if you have made changes before this step to overwrite the old list
#write.csv(unique_identifiers, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/identifiers.csv", row.names = FALSE)

```

# Data Cleaning, Part 2
## remove non-whale identifiers
```{r}

# Define a list of valid whale identifiers
valid_identifiers <- c("#723", "#49", "#383", "CRC-44", "CRC-56", "#44", "Shrimp", "Gray", "#185", "Shackleton", "#21", "Earhart", "#22","CRC-49", "CRC-53", "Saratoga Gray", "#724", "CRC-531", "CRC-383", "CRC-21", "Possible Gray", "#531", "CRC-185", "Patch", "Lucyfer", "Dubknuck", "Lucyfe", "#56", "Gray Whale", "Little Patch", "#53", "Shack", "Newbie", "Likely Shackleton", "Sound Grey", "Whale", "#2259", "Shackelton", "Dubnunk", "#2258", "#356", "Possibly Dubknuck", "CRC-2246", "Dubnuk", "Dubnuck", "Juv", "CRC-2262", "Lucy", "#2246", "Gray Whales", "Earheart", "Stardust", "Hattie", "Sounder", "Thidwick", "Gretchen", "Cascade", "Carson", "Tahoma", "Stalwart", "Eaerhart", "Gisborne", "CRC-723", "CRC-22", "CRC-2549", "Delta", "CRC-2234", "Crc", "Cretchen", "CRC-2255", "#2262", "Newb", "Adult Grays", "Lucifer", "#2249", "Patches", "#42", "#33", "#2356", "#2261", "#2253", "#2252", "#2251", "#2234", "PCFG 185", "CRC2246", "young gray whale", "patch", "little patch", "53", "49", "56", "352", "185", "723")

# Filter whale_identifiers to keep only valid ones
data <- data %>%
  mutate(whale_identifiers = lapply(whale_identifiers, function(ids) {
    ids[ids %in% valid_identifiers]
  }))

```

## re-check for unique identifiers
## list identified whales, unique names
```{r}

# we have lists of IDs in the comments column, but not all of them are actual IDs but rather observation note (i.e., young whale) and we want to know what we're working with. This code came from ChatGPT.

library(stringr)

# Define the regex pattern used for extraction
pattern <- "(#[\\d]+|CRC-\\d+|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"

# Extract identifiers from comments
extracted_identifiers <- str_extract_all(data$comments, pattern)

# Flatten the list and get unique values
unique_identifiers <- unique(unlist(extracted_identifiers))

# View the unique identifiers
print(unique_identifiers)

# write out to your data_adjusted output folder

# unblock this if you have made changes before this step to overwrite the old list
#write.csv(unique_identifiers, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/identifiers2.csv", row.names = FALSE)

```

## unify multiple identifiers for an individual
```{r}

# Create a synonym mapping (key-value pairs)
identifier_mapping <- c("A1" = "A", "A2" = "A", "B1" = "B", "B2" = "B")  # Example

# Standardize identifiers
data <- data %>%
  mutate(whale_identifiers = lapply(whale_identifiers, function(ids) {
    sapply(ids, function(id) ifelse(id %in% names(identifier_mapping), identifier_mapping[id], id))
  }))

```

## adjust the df to account for multiple whales listed in one comment line
```{r}

library(tidyr)

# Assume the `count` column represents the total number of whales observed
data <- data %>%
  unnest(whale_identifiers) %>%  # Each identifier gets its own row
  group_by(row_id = row_number()) %>%  # Add unique row identifier
  mutate(count = count / n()) %>%  # Divide count by number of identifiers
  ungroup() %>%
  select(-row_id)  # Remove helper column

```

# EDA
## total count 
```{r}

# total number of sounders sighted from 2013 - 2023
# save the table it creates if you don't want to copy paste the data OR you can knit this markdown file to a pdf or html file that supports printing/ manipulation.

total_count <- sum(data$count, na.rm = TRUE)
print(paste("Total Whale Count:", total_count))

```

## count by year
```{r}

yearly_counts <- data %>%
  group_by(year) %>%
  summarise(yearlyCount = sum(count, na.rm = TRUE))

# print
print(yearly_counts)

```


## count by month
```{r}

# total number of whales counted per month regardless of year

monthly_counts <- data %>%
  group_by(month) %>%
  summarise(totalCount = sum(count, na.rm = TRUE))

print(monthly_counts)

```

## count by season
```{r}

# group by season
seasonal_counts <- data %>%
  group_by(season) %>%
  summarise(totalCount = sum(count, na.rm = TRUE)) %>%
  arrange(match(season, c("Winter", "Spring", "Summer", "Fall")))

# print
print(seasonal_counts)

# plot seasonal trends
library(ggplot2)

ggplot(seasonal_counts, aes(x = season, y = totalCount, fill = season)) +
  geom_bar(stat = "identity") +
  labs(title = "Whale Sightings by Season",
       x = "Season",
       y = "Total Count of Sightings") +
  theme_minimal()

```

## count by season and year
```{r}

library(ggplot2)
library(dplyr)

# Group by year and season to get seasonal counts for each year
seasonal_counts_by_year <- data %>%
  group_by(year, season) %>%
  summarise(totalCount = sum(count, na.rm = TRUE)) %>%
  arrange(year, match(season, c("Winter", "Spring", "Summer", "Fall")))

# Print the grouped data
print(seasonal_counts_by_year)

# Plot seasonal trends by year
ggplot(seasonal_counts_by_year, aes(x = season, y = totalCount, fill = season)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ year) +  # Create a grid of plots, one for each year
  labs(title = "Whale Sightings by Season and Year",
       x = "Season",
       y = "Total Count of Sightings") +
  theme_minimal()

```


## count by quad
```{r}

# total number of whales counted per quad regardless of year, descending order
quad_counts <- data %>%
  group_by(quad) %>%
  summarise(totalCount = sum(count, na.rm = TRUE))%>%
  arrange(desc(totalCount))

print(quad_counts)

```


## 'map' of sightings
```{r}

# plotting with lat/ long coordinates, but not overlayed on a map - just want to see what it looks like

library(ggplot2)

ggplot(timeline_data, aes(x = long, y = lat, color = whale_identifiers)) +
  geom_point(alpha = 0.6) +
  labs(title = "Map of Whale Sightings", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```

## timeline of sightings
```{r}

# convert sightDate to Date format (if it wasn't done earlier)
#timeline_data$sightDate <- as.Date(timeline_data$sightDate, format = "%m/%d/%y")

ggplot(timeline_data, aes(x = sightDate, y = whale_identifiers)) +
  geom_point(aes(color = whale_identifiers), alpha = 0.6) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Timeline of Whale Sightings", x = "Date", y = "Whale Identifiers", color = "Whale Identifiers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## movement patterns
```{r}

# Sort by identifiers and date
timeline_data <- timeline_data[order(timeline_data$whale_identifiers, timeline_data$sightDate), ]

ggplot(timeline_data, aes(x = long, y = lat, group = whale_identifiers, color = whale_identifiers)) +
  geom_path(alpha = 0.6) +
  geom_point(alpha = 0.6) +
  labs(title = "Movement Patterns of Identifiable Whales", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```


