---
title: EDA 00.2: All Whales w/ Sounder designator Data Prep & Exploration
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data

## working with all whales

```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/all_whales_with_sounder_ids_cleaned.csv")

```

# Counts, overall

## Sounder Designation Keys

Whale Identifiers:\
none= whales with no comments to tell us who they are\
unknown= comment said Sounder but no ID confirmed or reporter was unsure\
ID= an identified whale that may or may not be a Sounder.\

Sounder column:\
unknown= comment said Sounder but no ID confirmed\
n= a whale with or without an ID that isn't a Sounder\
y= an identified Sounder

### total, complete count

```{r}

# total number of whales sighted from 2013 - 2023 = 8872

total_count <- sum(data$count, na.rm = TRUE)
print(paste("Total Whale Count:", total_count))

```

### total, count breakdown by Sounder status

```{r}

# total number breakdown: 2327 Sounders, 293 unknown, 6252 non-Sounders

# group by Sounder's designation
counts_by_sounder <- aggregate(count ~ sounders, data = data, sum, na.rm = TRUE)

# print
print("Total Whale Count by Sounder:")
print(counts_by_sounder)

```

### total, count breakdown by whale identifier

```{r}

# 22 and 53 (Earhart & Little Patch are seen the most); remaining data in the table below

#group by identifier
counts_by_whale <- aggregate(count ~ whale_identifier, data = data, sum, na.rm = TRUE)

# print 
print("Total Whale Count by Whale Identifier:")
print(counts_by_whale)

```

### counts, excluding unidentified whales

```{r}

# histogram of sightings, just to help visualize the table above (25 of the 58 identified whales were only seen once... worth investigation if there's time)

library(dplyr)
library(ggplot2)

# filter out 'none' and 'unknown' whales
filtered_data <- data %>%
  filter(!whale_identifier %in% c("none", "unknown"))

# group by whale_identifier and count sightings
sightings_per_whale <- filtered_data %>%
  group_by(whale_identifier) %>%
  summarise(sightings_count = sum(count, na.rm = TRUE)) %>%
  arrange(desc(sightings_count))

# plot a histogram 
ggplot(sightings_per_whale, aes(x = sightings_count)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Whale Sightings Count",
       x = "Number of Sightings",
       y = "Number of Whales") +
  theme_minimal()

```

### proportion of sightings under 10 per whale

```{r}

# calculate proportion of whales with sightings under 10 = 58.62%. More than half of the ID'd whales were seen less than 10 times in 10 years.

proportion_under_10 <- sightings_per_whale %>%
  summarise(proportion = mean(sightings_count < 10)) %>%
  pull(proportion)

print(paste("Proportion of whales with sightings under 10:", round(proportion_under_10 * 100, 2), "%"))

```

### sightings stats

```{r}

# compute basic stats on sightings count
stats <- sightings_per_whale %>%
  summarise(
    mean_count = mean(sightings_count),
    median_count = median(sightings_count),
    max_count = max(sightings_count),
    min_count = min(sightings_count),
    q1 = quantile(sightings_count, 0.25),
    q3 = quantile(sightings_count, 0.75)
  )

print("Summary Statistics for Sightings Count:")
print(stats)

# whales sightings in each quartile
quartiles <- sightings_per_whale %>%
  mutate(
    quartile = case_when(
      sightings_count <= stats$q1 ~ "Q1",
      sightings_count > stats$q1 & sightings_count <= stats$median_count ~ "Q2",
      sightings_count > stats$median_count & sightings_count <= stats$q3 ~ "Q3",
      sightings_count > stats$q3 ~ "Q4"
    )
  )

print("Whales by Quartile:")
print(quartiles)

```

## Sounder confirmed & unknown sighting plot by whale for 10y span

```{r}

library(dplyr)
library(ggplot2)
library(tidyr)

# filter data to include only sounders ('y' or 'unknown')
filtered_data <- data %>%
  filter(sounders %in% c("y", "unknown"))

# group by year and whale_identifier & count the number of sounders identified per year
sounders_by_year <- filtered_data %>%
  group_by(year, whale_identifier) %>%
  summarise(sounder_count = n(), .groups = "drop")

# generate plots for each whale spanning from 2013 to 2023
unique_whales <- unique(sounders_by_year$whale_identifier)

for (whale in unique_whales) {
  whale_data <- sounders_by_year %>%
    filter(whale_identifier == whale) %>%
    complete(year = 2013:2023, fill = list(sounder_count = 0)) # Fill missing years with 0 counts
  
  p <- ggplot(whale_data, aes(x = year, y = sounder_count)) +
    geom_bar(stat = "identity", fill = "blue", color = "black", alpha = 0.7) +
    labs(
      title = paste("Sounders Identified Per Year for Whale:", whale),
      x = "Year",
      y = "Sounder Count"
    ) +
    theme_minimal() +
    scale_x_continuous(breaks = seq(2013, 2023, 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
  
  # Save each plot to a file
  #ggsave(filename = paste0("sounder_plot_", whale, ".png"), plot = p, width = 8, height = 6)
}

```

# Annual plots

## number of sounders sightings by whale

```{r}

library(dplyr)
library(ggplot2)
library(tidyr)

# Filter data to include only sounders ('y' or 'unknown')
filtered_data <- data %>%
  filter(sounders %in% c("y", "unknown"))

# Group by year and whale_identifier and count the number of sounders identified per year
sounders_by_year <- filtered_data %>%
  group_by(year, whale_identifier) %>%
  summarise(sounder_count = n(), .groups = "drop")

# Generate plots for each year
unique_years <- unique(sounders_by_year$year)

for (year in unique_years) {
  year_data <- sounders_by_year %>%
    filter(year == !!year) %>%
    complete(whale_identifier = unique(filtered_data$whale_identifier), fill = list(sounder_count = 0)) # Fill missing whales with 0 counts
  
  p <- ggplot(year_data, aes(x = reorder(whale_identifier, -sounder_count), y = sounder_count)) +
    geom_bar(stat = "identity", fill = "blue", color = "black", alpha = 0.7) +
    labs(
      title = paste("Sounders Identified for All Whales in Year:", year),
      x = "Whale Identifier",
      y = "Sounder Count"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
  
  # Save each plot to a file
  #ggsave(filename = paste0("sounder_plot_year_", year, ".png"), plot = p, width = 10, height = 6)
}


```

## table of sightings by year, cumulative sounder counts

```{r}

# Group data by year and sounder and count sightings
sightings_by_year <- filtered_data %>%
  group_by(year, sounders) %>%
  summarise(sightings_count = n(), .groups = "drop")

# View the table
print(sightings_by_year)


```

## line plot of sightings by year - this is in eda 00.1 as well

```{r}

library(ggplot2)

# Line plot of sightings by year
ggplot(sightings_by_year, aes(x = year, y = sightings_count, color = sounders, group = sounders)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Sightings by Year",
    x = "Year",
    y = "Number of Sightings",
    color = "Sounder"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(sightings_by_year$year), max(sightings_by_year$year), 1))


```

## stacked bar of sightings by year

### this can be cleaned up, and re-ordered. i think this is great for an overview going into the Sounders data

```{r}

# Ensure data includes whale_identifier counts by year
sightings_by_year_whale <- filtered_data %>%
  group_by(year, whale_identifier) %>%
  summarise(sightings_count = n(), .groups = "drop")

# Stacked bar plot color-coded by individual whale
ggplot(sightings_by_year_whale, aes(x = year, y = sightings_count, fill = whale_identifier)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Sightings by Year (Stacked by Whale Identifier)",
    x = "Year",
    y = "Number of Sightings",
    fill = "Whale Identifier"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(sightings_by_year_whale$year), max(sightings_by_year_whale$year), 1)) +
  theme(legend.position = "right")  # Adjust legend placement


```

## table of sightings by month regardless of year, cumulative sounder counts

```{r}

# Group data by month and sounder and count sightings
sightings_by_month <- filtered_data %>%
  group_by(month, sounders) %>%
  summarise(sightings_count = n(), .groups = "drop") %>%
  arrange(as.numeric(month)) # Arrange months in numeric order

# View the table
print(sightings_by_month)


```
