---
title: EDA 00.3: Reporting Channel and Source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
## working with all whales
## NOTE: Sightings data is from January 1, 2013 - December 31, 2023 but data recorded from TMW ends at April 1, 2023.
```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/all_whales_with_sounder_ids_cleaned.csv")

```

# Sightings Source/ Reporters
## channel
```{r}

# filter out 'none' and 'unknown' whales
#filtered_data <- data %>%
  #filter(!whale_identifier %in% c("none", "unknown"))

# Summarize counts for each channel
channel_summary <- data %>%
  group_by(channel) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percentage = (count / sum(count)) * 100)

# Update channel_summary to include percentages in the legend labels with capitalized channel names
channel_summary <- channel_summary %>%
  mutate(
    channel = str_to_title(channel), # Capitalize the first letter of each channel name
    legend_label = paste0(channel, " (", round(percentage, 1), "%)")
  )

# Pie chart with percentages in the legend and larger text
channel <- ggplot(channel_summary, aes(x = "", y = count, fill = legend_label)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Reporting Channels",
    fill = "Channel (with Percentage)"
  ) +
  theme_void() +
  theme(
    legend.title = element_text(size = 14), # Increase legend title font size
    legend.text = element_text(size = 12), # Increase legend text font size
    plot.title = element_text(size = 16, hjust = 0.5) # Center and enlarge title
  )

# Save the plot
ggsave(filename = "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/channel_pie.png",
       plot = channel, width = 12, height = 10, dpi = 300)

# Print the plot
print(channel)

```

## source
```{r}

library(dplyr)
library(ggplot2)
library(stringr)

# Define categories
source_categories <- c("public", "reliable")

# Summarize counts for each source, categorize, and aggregate "Unknown" sources
source_summary <- data %>%
  group_by(source) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    percentage = (count / sum(count)) * 100,
    category = case_when(
      source %in% source_categories ~ source,  # Assign as "Public" or "Reliable"
      TRUE ~ "Unknown"                         # Assign as "Unknown" if not in categories
    )
  ) %>%
  group_by(category) %>%  # Aggregate counts for the "Unknown" category
  summarise(
    count = sum(count),
    percentage = sum(percentage),
    .groups = "drop"
  )

# Capitalize category names and create legend labels
source_summary <- source_summary %>%
  mutate(
    category = str_to_title(category), # Capitalize the category names
    legend_label = paste0(category, " (", round(percentage, 1), "%)")
  )

# Pie chart with percentages in the legend
source <- ggplot(source_summary, aes(x = "", y = count, fill = legend_label)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Reporting Sources",
    fill = "Source (with Percentage)"
  ) +
  theme_void() +
  theme(
    legend.title = element_text(size = 14),  # Larger legend title
    legend.text = element_text(size = 12),  # Larger legend text
    plot.title = element_text(size = 16, hjust = 0.5)  # Center and enlarge title
  )

# Save the plot
ggsave(filename = "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/source_pie.png",
       plot = source, width = 12, height = 10, dpi = 300)

# Print the plot
print(source)


```

# Number of whale sightings by sounder v not sounder designation
```{r}

library(dplyr)
library(ggplot2)

# Summarize counts for each Sounders category (Y and N)
sightings_summary <- data %>%
  group_by(sounders) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  mutate(
    percentage = (count / sum(count)) * 100,
    category = case_when(
      sounders == "y" ~ "Sounders",
      TRUE ~ "Other/Unknown"
    )
  ) %>%
  group_by(category) %>%  # Aggregate counts for categories
  summarise(
    count = sum(count),
    percentage = sum(percentage),
    .groups = "drop"
  ) %>%
  mutate(
    legend_label = paste0(category, " (", round(percentage, 1), "%)")
  )

# Create the pie chart
whale_chart <- ggplot(sightings_summary, aes(x = "", y = count, fill = legend_label)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Total Gray Whale Sightings 2013 - 2023",
    fill = "Whales"
  ) +
  theme_void() +
  theme(
    legend.title = element_text(size = 14),  # Larger legend title
    legend.text = element_text(size = 12),  # Larger legend text
    plot.title = element_text(size = 16, hjust = 0.5)  # Center and enlarge title
  )

# Save the chart
ggsave(filename = "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/whale_pie.png",
       plot = whale_chart, width = 12, height = 10, dpi = 300)

# Print the chart
print(whale_chart)


```


# Temporal Scale
## monthly breakdown of reliable v public sightings regardless of year
```{r}

# Summarize sightings by month and source
monthly_summary <- filtered_data %>%
  group_by(month, source) %>%
  summarise(sightings_count = n(), .groups = "drop")

# Stacked bar plot: Reliable vs Public by month over 10 years
annual_source<- ggplot(monthly_summary, aes(x = month, y = sightings_count, fill = source)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Reliable vs Public Sightings (Month by Month) Over 10 Years",
    x = "Month",
    y = "Number of Sightings",
    fill = "Source"
  ) +
  scale_x_discrete(limits = sprintf("%02d", 1:12), labels = month.abb) +  # Use month abbreviations
  theme_minimal()

ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/annual_source.png", plot=annual_source, width = 12, height = 10, dpi = 300)

```

## sightings by year, reliable v public
```{r}


library(dplyr)
library(ggplot2)
library(stringr)

# Define categories
source_categories <- c("public", "reliable")

# Summarize counts for each source and categorize
data <- data %>%
  mutate(
    category = case_when(
      source %in% source_categories ~ str_to_title(source),  # Assign as "Public" or "Reliable" (capitalize names)
      TRUE ~ "Unknown"  # Assign as "Unknown" if not in categories
    )
  )

# Summarize sightings by year, month, and category
yearly_summary <- data %>%
  group_by(year, month, category) %>%
  summarise(sightings_count = n(), .groups = "drop") %>%
  mutate(
    month = sprintf("%02d", as.integer(month))  # Ensure month format is zero-padded for sorting
  )

# Facet-wrapped stacked bar plots
yearly_source2 <- ggplot(yearly_summary, aes(x = month, y = sightings_count, fill = category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Reliable vs Public Sightings by Year",
    x = "Month",
    y = "Number of Sightings",
    fill = "Source"
  ) +
  scale_x_discrete(limits = sprintf("%02d", 1:12), labels = month.abb) +  # Use month abbreviations
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),  # Increase facet label size
    axis.text.x = element_text(angle = 45, hjust = 1)  # Adjust month axis text
  ) +
  facet_wrap(~ year, ncol = 3)  # Adjust columns as needed

# Print the plot
print(yearly_source2)

# Save the plot
ggsave(filename = "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/yearly_source2.png", plot = yearly_source2, width = 12, height = 10, dpi = 300)

```

## time of day breakdown
## Time categories:\
Not recorded= 0:00 or 9:99
Morning= times between 0:01 - 9:59
Midday= times between 10:00 - 16:59
Evening= times between 17:00 - 20:59
Night- times between 21:00 - 23:59
```{r}

library(dplyr)

# Convert time to POSIXct for proper comparison
data <- data %>%
  mutate(
    time_clean = ifelse(time == "9:99" | time == "0:00", NA, paste0(time, ":00")), # Fix missing seconds
    time_parsed = as.POSIXct(time_clean, format = "%H:%M:%S", tz = "PST"),
    time_of_day = case_when(
      is.na(time_parsed) ~ "Not recorded",                       # Handle NA times
      time_parsed >= as.POSIXct("00:00:01", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("03:59:59", format = "%H:%M:%S") ~ "Before Sunrise",  # 0:01 to 3:59
      time_parsed >= as.POSIXct("04:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("05:59:59", format = "%H:%M:%S") ~ "Early Morning",  # 4:00 to 5:59
      time_parsed >= as.POSIXct("06:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("09:59:59", format = "%H:%M:%S") ~ "Late Morning",  # 6:00 to 9:59
      time_parsed >= as.POSIXct("10:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("16:59:59", format = "%H:%M:%S") ~ "Midday",   # 10:00 to 16:59
      time_parsed >= as.POSIXct("17:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("20:59:59", format = "%H:%M:%S") ~ "Evening",  # 17:00 to 20:59
      time_parsed >= as.POSIXct("21:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("23:59:59", format = "%H:%M:%S") ~ "Night",    # 21:00 to 23:59
      TRUE ~ "Not recorded"                                    # Catch any remaining cases
    )
  ) %>%
  select(-time_clean, -time_parsed)  # Drop intermediate columns if no longer needed

# View the updated dataset
head(data)


```

## troubleshoot time problem
```{r}

head(data$time)

```


## counts by time of day
### based on the general results below, I'm going to break morning into two groups to clarify since the results are as follows\
Evening & Night = 0 sightings\
Morning= 5324\
Midday= 551\
Not recorded= 1201\
```{r}

# the shift to early v late morning breakdown shows an interesting trend; most sightings are between 12:01am - 5:59am
library(tidyr)

# Define all possible time_of_day categories
all_time_of_day <- c("Early Morning", "Late Morning", "Midday", "Evening", "Night", "Not recorded")

# Summarize total sightings by time_of_day, including times with 0 sightings
sightings_by_time_of_day <- data %>%
  group_by(time_of_day) %>%
  summarise(total_sightings = n(), .groups = "drop") %>%
  complete(time_of_day = all_time_of_day, fill = list(total_sightings = 0))

# View the summary
print(sightings_by_time_of_day)


```

# sightings by location
```{r}

library(ggplot2)
library(sf)
library(dplyr)
library(ggspatial)
library(maps)

# Filter the data for Sounders only
sounder_whales <- data %>%
  filter(sounders == "y")

# Convert data to an sf object for mapping
sounder_sf <- st_as_sf(sounder_whales, coords = c("long", "lat"), crs = 4326)

# Load a base map and focus on Washington State
usa <- st_as_sf(map("state", plot = FALSE, fill = TRUE))  # US states
washington <- usa %>%
  filter(ID == "washington")  # Filter for Washington State

# Plot the sightings on the Washington map
ggplot() +
  geom_sf(data = washington, fill = "lightgrey", color = "white") +
  geom_sf(data = sounder_sf, aes(color = whale_identifier), size = 2, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +  # Use a color palette suitable for categorical data
  labs(
    title = "Sounder Whale Sightings in Washington State by Identifier",
    color = "Whale Identifier",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

```

# zoom in
```{r}

library(ggplot2)
library(sf)
library(dplyr)
library(maps)


# Filter the data for Sounders only
sounder_whales <- data %>%
  filter(sounders == "y")

# Convert data to an sf object for mapping
sounder_sf <- st_as_sf(sounder_whales, coords = c("long", "lat"), crs = 4326)

# Load a base map and focus on Washington State
usa <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
washington <- usa %>%
  filter(ID == "washington")

# Define the coordinate limits to zoom into a specific area
# Replace these with the desired longitude and latitude ranges
x_limits <- c(-123.5, -122)  # Longitude range
y_limits <- c(47, 49)      # Latitude range

# Plot the sightings on the Washington map with zoomed-in area
ggplot() +
  geom_sf(data = washington, fill = "lightgrey", color = "white") +
  geom_sf(data = sounder_sf, aes(color = whale_identifier), size = 3, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo") +
  coord_sf(xlim = x_limits, ylim = y_limits, expand = FALSE) +
  labs(
    title = "Sounder Whale Sightings Zoomed In by Identifier",
    color = "Whale Identifier",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )


```

