---
title: EDA 00.3: Reporting Channel and Source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
## working with all whales
## NOTE: Sightings data is from January 1, 2013 - December 31, 2023 but data recorded from TMW ends at April 1, 2023.
```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/all_whales_with_sounder_ids_cleaned.csv")

```

# Sightings Source/ Reporters
## channel
```{r}

# filter out 'none' and 'unknown' whales
filtered_data <- data %>%
  filter(!whale_identifier %in% c("none", "unknown"))

# Summarize counts for each channel
channel_summary <- filtered_data %>%
  group_by(channel) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percentage = (count / sum(count)) * 100)

# Update channel_summary to include percentages in the legend labels
channel_summary <- channel_summary %>%
  mutate(legend_label = paste0(channel, " (", round(percentage, 1), "%)"))

# Pie chart with percentages in the legend
channel<- ggplot(channel_summary, aes(x = "", y = count, fill = legend_label)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Distribution of Channels",
    fill = "Channel (with Percentage)"
  ) +
  theme_void()

ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/channel_pie.png", plot=channel, width = 12, height = 10, dpi = 300)

print(channel)
```

## source
```{r}

# Summarize counts for each source
source_summary <- filtered_data %>%
  group_by(source) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percentage = (count / sum(count)) * 100)

# Update source_summary to include percentages in the legend labels
source_summary <- source_summary %>%
  mutate(legend_label = paste0(source, " (", round(percentage, 1), "%)"))

# Pie chart with percentages in the legend for source
source<- ggplot(source_summary, aes(x = "", y = count, fill = legend_label)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Distribution of Sources",
    fill = "Source (with Percentage)"
  ) +
  theme_void()

ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/source_pie.png", plot=source, width = 12, height = 10, dpi = 300)

```

# Temporal Scale
## monthly breakdown of reliable v public sightings regardless of year
```{r}

# Summarize sightings by month and source
monthly_summary <- filtered_data %>%
  group_by(month, source) %>%
  summarise(sightings_count = n(), .groups = "drop")

# Stacked bar plot: Reliable vs Public by month over 10 years
annual_source<- ggplot(monthly_summary, aes(x = month, y = sightings_count, fill = source)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Reliable vs Public Sightings (Month by Month) Over 10 Years",
    x = "Month",
    y = "Number of Sightings",
    fill = "Source"
  ) +
  scale_x_discrete(limits = sprintf("%02d", 1:12), labels = month.abb) +  # Use month abbreviations
  theme_minimal()

ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/annual_source.png", plot=annual_source, width = 12, height = 10, dpi = 300)

```

## sightings by year, reliable v public
```{r}

# Summarize sightings by year, month, and source
yearly_summary <- filtered_data %>%
  group_by(year, month, source) %>%
  summarise(sightings_count = n(), .groups = "drop")

# Facet-wrapped stacked bar plots
yearly_source<- ggplot(yearly_summary, aes(x = month, y = sightings_count, fill = source)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Reliable vs Public Sightings by Year",
    x = "Month",
    y = "Number of Sightings",
    fill = "Source"
  ) +
  scale_x_discrete(limits = sprintf("%02d", 1:12), labels = month.abb) +  # Use month abbreviations
  theme_minimal() +
  facet_wrap(~ year, ncol = 3)  # Adjust columns as needed

ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_dp/output/figures/yearly_source.png", plot=yearly_source, width = 12, height = 10, dpi = 300)

```

## time of day breakdown
## Time categories:\
Not recorded= 0:00 or 9:99
Morning= times between 0:01 - 9:59
Midday= times between 10:00 - 16:59
Evening= times between 17:00 - 20:59
Night- times between 21:00 - 23:59
```{r}

library(dplyr)

# Convert time to POSIXct for proper comparison
data <- data %>%
  mutate(
    time_clean = ifelse(time == "9:99" | time == "0:00", NA, paste0(time, ":00")), # Fix missing seconds
    time_parsed = as.POSIXct(time_clean, format = "%H:%M:%S", tz = "PST"),
    time_of_day = case_when(
      is.na(time_parsed) ~ "Not recorded",                       # Handle NA times
      time_parsed >= as.POSIXct("00:00:01", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("03:59:59", format = "%H:%M:%S") ~ "Before Sunrise",  # 0:01 to 3:59
      time_parsed >= as.POSIXct("04:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("05:59:59", format = "%H:%M:%S") ~ "Early Morning",  # 4:00 to 5:59
      time_parsed >= as.POSIXct("06:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("09:59:59", format = "%H:%M:%S") ~ "Late Morning",  # 6:00 to 9:59
      time_parsed >= as.POSIXct("10:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("16:59:59", format = "%H:%M:%S") ~ "Midday",   # 10:00 to 16:59
      time_parsed >= as.POSIXct("17:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("20:59:59", format = "%H:%M:%S") ~ "Evening",  # 17:00 to 20:59
      time_parsed >= as.POSIXct("21:00:00", format = "%H:%M:%S") & 
      time_parsed <= as.POSIXct("23:59:59", format = "%H:%M:%S") ~ "Night",    # 21:00 to 23:59
      TRUE ~ "Not recorded"                                    # Catch any remaining cases
    )
  ) %>%
  select(-time_clean, -time_parsed)  # Drop intermediate columns if no longer needed

# View the updated dataset
head(data)


```

## troubleshoot time problem
```{r}

head(data$time)

```


## counts by time of day
### based on the general results below, I'm going to break morning into two groups to clarify since the results are as follows\
Evening & Night = 0 sightings\
Morning= 5324\
Midday= 551\
Not recorded= 1201\
```{r}

# the shift to early v late morning breakdown shows an interesting trend; most sightings are between 12:01am - 5:59am
library(tidyr)

# Define all possible time_of_day categories
all_time_of_day <- c("Early Morning", "Late Morning", "Midday", "Evening", "Night", "Not recorded")

# Summarize total sightings by time_of_day, including times with 0 sightings
sightings_by_time_of_day <- data %>%
  group_by(time_of_day) %>%
  summarise(total_sightings = n(), .groups = "drop") %>%
  complete(time_of_day = all_time_of_day, fill = list(total_sightings = 0))

# View the summary
print(sightings_by_time_of_day)


```


