---
title: Plots 01: Sounders' Movements and Habits
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
## only working with the Sounders' data in this markdown file

```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_added_rows.csv")

```

# Temporal Visualization
## line chart for sightings by year
### clean up the labels manually using powerpoint/ google slides and this should go into your presentation. Essentially remove the labels and title by covering them with white blocks; make the title of the table the slide title, and replace the text with easier to read font/ size. group the image & changes together so that nothing moves out of place if you adjust it.
```{r}

library(ggplot2)

# make sure your years, months, quads are being treated like characters, not numbers
data$year <- as.character(data$year)
data$month <- as.character(data$month)
data$quad <- as.character(data$quad)

# summarize sightings by year
sightings_by_year <- data %>%
  group_by(year) %>%
  summarize(total_sightings = sum(count))

# create a line chart 
ggplot(sightings_by_year, aes(x = year, y = total_sightings, group = 1)) +
  geom_line() +
  geom_point() +
  ggtitle("Sightings Count Over Years") +
  xlab("Year") +
  ylab("Total Sightings") +
  theme_minimal() +
  scale_x_discrete() 

```

## heatmap of sightings by month and year
### interesting but not presentation- level interesting; this confirms what we see in the data, so that makes it a nice check 
```{r}

# make `month` a factor with levels 1 to 12
data$month <- factor(data$month, levels = as.character(1:12))

# summarize sightings by month and year
sightings_heatmap <- data %>%
  group_by(year, month) %>%
  summarize(total_sightings = sum(count))

# create the heatmap
ggplot(sightings_heatmap, aes(x = month, y = factor(year), fill = total_sightings)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Sightings") +
  ggtitle("Heatmap of Sightings by Month and Year") +
  xlab("Month") +
  ylab("Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

## polar chart sightings by season
### I thought this could be fun - i wasn't expecting the spring to sooooo heavily skew the plot
```{r}

# Summarize sightings by season
sightings_by_season <- data %>%
  group_by(season) %>%
  summarize(total_sightings = sum(count))

# Create a polar chart
ggplot(sightings_by_season, aes(x = season, y = total_sightings, fill = season)) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "x") +
  ggtitle("Sightings by Season") +
  xlab("") +
  ylab("Total Sightings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## count by season and year
### I'm a sucker for a wrapped plot. It's the full 10 years in a single view - too much for a presenation, but can be split up.
```{r}

library(ggplot2)
library(dplyr)

# Group by year and season to get seasonal counts for each year
seasonal_counts_by_year <- data %>%
  group_by(year, season) %>%
  summarise(totalCount = sum(count, na.rm = TRUE)) %>%
  arrange(year, match(season, c("Winter", "Spring", "Summer", "Fall")))

# Print the grouped data
print(seasonal_counts_by_year)

# Plot seasonal trends by year
ggplot(seasonal_counts_by_year, aes(x = season, y = totalCount, fill = season)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ year) +  # Create a grid of plots, one for each year
  labs(title = "Whale Sightings by Season and Year",
       x = "Season",
       y = "Total Count of Sightings") +
  theme_minimal()

```

## seasonality for top 10 whales
### check out 2440 - this whale is interesting and doesn't appear to leave the area for long at all.
```{r}

# Summarize sightings of each whale by season
whale_seasonal_trends <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier, season) %>%
  summarize(total_sightings = sum(count), .groups = "drop")

# Select top whales 
top_whales <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier) %>%
  summarize(total_sightings = sum(count), .groups = "drop") %>%
  arrange(desc(total_sightings)) %>%
  slice_head(n = 10)

# Filter seasonal trends for top whales
top_whale_seasonal <- whale_seasonal_trends %>%
  filter(whale_identifier %in% top_whales$whale_identifier)

# Create a faceted plot for seasonality
ggplot(top_whale_seasonal, aes(x = season, y = total_sightings, fill = season)) +
  geom_col() +
  facet_wrap(~whale_identifier, scales = "free_y") +
  ggtitle("Seasonal Sightings for Top Whales") +
  xlab("Season") +
  ylab("Total Sightings") +
  theme_minimal() +
  theme(
    legend.position = "none", # Remove legend for simplicity
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )

```

## monthly trends for the top 10
### again - check out 2440. this whale is only 'gone' in october and november...
```{r}

# Summarize sightings of each whale by month
whale_monthly_trends <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier, month) %>%
  summarize(total_sightings = sum(count), .groups = "drop")

# Select top 10 whales by total sightings
top_whales <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier) %>%
  summarize(total_sightings = sum(count), .groups = "drop") %>%
  arrange(desc(total_sightings)) %>%
  slice_head(n = 10)

# Filter monthly trends for top whales
top_whale_monthly <- whale_monthly_trends %>%
  filter(whale_identifier %in% top_whales$whale_identifier)

# Ensure months are ordered correctly
top_whale_monthly$month <- factor(top_whale_monthly$month, levels = as.character(1:12))

# Create a faceted plot for monthly trends
ggplot(top_whale_monthly, aes(x = month, y = total_sightings, fill = month)) +
  geom_col() +
  facet_wrap(~whale_identifier, scales = "free_y") +
  ggtitle("Monthly Sightings for Top 10 Whales") +
  xlab("Month") +
  ylab("Total Sightings") +
  theme_minimal() +
  theme(
    legend.position = "none", # Remove legend for simplicity
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )

```


# Geographic Visualization
## scatterplot base for future mapping
```{r}

library(ggplot2)

# Scatter plot for geographic distribution
ggplot(data, aes(x = long, y = lat)) +
  geom_point(alpha = 0.5, size = 2) +
  ggtitle("Geographic Distribution of Sightings") +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_minimal()

```

## density of sightings' locations (confirms the scatterplot locations above and can be overlayed on a map)
```{r}

# density plot for sightings concentration
ggplot(data, aes(x = long, y = lat)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", alpha = 0.5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  ggtitle("Density of Sightings by Location") +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_minimal()

```

# Individual Whale Trends
## top 10 most seen
```{r}

top_whales <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier) %>%
  summarize(total_sightings = sum(count)) %>%
  arrange(desc(total_sightings)) %>%
  head(10)

ggplot(top_whales, aes(x = reorder(whale_identifier, total_sightings), y = total_sightings)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  ggtitle("Top 10 Most Sighted Whales") +
  xlab("Whale Identifier") +
  ylab("Total Sightings") +
  theme_minimal()

```

## known v unknown (unconfirmed or missing identifier)
### interesting to know almost 90% of sounder IDs are correct and corraborated. We know from the EDA work that there are 58 unique whales and the category of 'unknown' for a total of 59 unique ID catefories.
```{r}

library(ggplot2)

# proportion of unique whales vs unknown
whale_proportions <- data %>%
  mutate(category = if_else(whale_identifier == "unknown", "Unknown", "Identified")) %>%
  group_by(category) %>%
  summarize(total_sightings = sum(count))

# percentages
whale_proportions <- whale_proportions %>%
  mutate(percentage = total_sightings / sum(total_sightings) * 100)

# Create a pie chart with percentages
ggplot(whale_proportions, aes(x = "", y = total_sightings, fill = category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette = "Set2") +
  ggtitle("Proportion of Unique Whales vs. Unknown") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),  # Remove X-axis title
        axis.title.y = element_blank(),  # Remove Y-axis title
        axis.text = element_blank(),     # Remove axis text
        axis.ticks = element_blank())    # Remove axis ticks


```

## sightings over time for each ID'd whale (change the plot code to whichever identifier you want)
```{r}

library(ggplot2)

data$year <- as.character(data$year)

# summarize sightings of each whale by year
whale_yearly_trends <- data %>%
  filter(whale_identifier != "unknown") %>% # exclude unknown whales
  group_by(whale_identifier, year) %>%
  summarize(total_sightings = sum(count), .groups = "drop")

# Plot for a single whale (e.g., whale "383")
ggplot(whale_yearly_trends %>% filter(whale_identifier == "383"),
       aes(x = year, y = total_sightings, group = 1)) +
  geom_line() +
  geom_point() +
  ggtitle("Sightings Over Time for Whale 383") +
  xlab("Year") +
  ylab("Total Sightings") +
  theme_minimal() +
  scale_x_discrete() 


```

## compare whales
### expand this visual out and let me know what you want to see out of it - it is very interesting.
```{r}

# top 10 most sighted whales
top_whales <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier) %>%
  summarize(total_sightings = sum(count), .groups = "drop") %>%
  arrange(desc(total_sightings)) %>%
  slice_head(n = 10)

# Filter yearly trends for top whales
top_whale_trends <- whale_yearly_trends %>%
  filter(whale_identifier %in% top_whales$whale_identifier)

# Plot trends for top whales using facets
ggplot(top_whale_trends, aes(x = year, y = total_sightings, color = whale_identifier, group = 1)) +
  geom_line() +
  geom_point() +
  facet_wrap(~whale_identifier, scales = "free_y") +
  ggtitle("Sightings Over Time for Top  10 Whales") +
  xlab("Year") +
  ylab("Total Sightings") +
  theme_minimal()+
  scale_x_discrete()


```

## plotting the same data as above on a single plot
```{r}

# top 10 most sighted whales
top_whales <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(whale_identifier) %>%
  summarize(total_sightings = sum(count), .groups = "drop") %>%
  arrange(desc(total_sightings)) %>%
  slice_head(n = 10)

# Filter yearly trends for top whales
top_whale_trends <- whale_yearly_trends %>%
  filter(whale_identifier %in% top_whales$whale_identifier)

# Plot all top whales on the same chart
ggplot(top_whale_trends, aes(x = year, y = total_sightings, color = whale_identifier, group = whale_identifier, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  ggtitle("Sightings Over Time for Top Whales") +
  xlab("Year") +
  ylab("Total Sightings") +
  theme_minimal() +
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10)) +
  scale_color_discrete(name = "Whale Identifier") + # Custom legend title
  scale_x_discrete()



```



# Group Trends
## distribution of group sizes
## Are there groups that travel/ feed/ whatever together or do they operate solo?
### dominantly - they are seen alone with about half as many sightings seeing an identified pair of whales.
```{r}

# count unique whales per group
group_composition <- data %>%
  group_by(sightDate, time, lat, long) %>% # Group by unique sighting event
  summarize(unique_whales = n_distinct(whale_identifier)) %>%
  ungroup()

# Plot group composition distribution
ggplot(group_composition, aes(x = unique_whales)) +
  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.7) +
  ggtitle("Distribution of Unique Whales per Group") +
  xlab("Number of Unique Whales") +
  ylab("Frequency") +
  theme_minimal()


```

## group cohesion
## How does group composition change over time? (could also be in temporal viz)
### no real change over time - group size leans heavily to 1 and obviously we cannot have .10 or .50 of a living whale
```{r}

data$year <- as.character(data$year)

# Group size by year
group_size_year <- data %>%
  group_by(year) %>%
  summarize(mean_group_size = mean(count))

ggplot(group_size_year, aes(x = year, y = mean_group_size, group = 1)) +
  geom_line() +
  geom_point() +
  ggtitle("Average Group Size by Year") +
  xlab("Year") +
  ylab("Average Group Size") +
  theme_minimal() +
  scale_x_discrete()

```


## What groups do we see when there are more than one traveling together?
### this gives us something far too busy... let's downsize and look at what we can see
```{r}

#install.packages('widyr')
library(widyr)
library(dplyr)
library(tidyr)
library(igraph)

# Prepare co-occurrence data frame step by step
grouped_data <- data %>%
  filter(whale_identifier != "unknown") %>%
  group_by(sightDate, time, lat, long) %>%
  summarize(whale_list = list(whale_identifier), .groups = "drop") %>%
  unnest(whale_list)

# Rename the column explicitly to match `pairwise_count` requirements
grouped_data <- grouped_data %>%
  rename(item = whale_list) # Rename to `item` for clarity

# Manual co-occurrence calculation
co_occurrence <- grouped_data %>%
  group_by(item) %>%
  mutate(dummy = 1) %>%
  tidyr::crossing(item2 = item) %>%
  filter(item != item2) %>%
  group_by(item, item2) %>%
  summarize(count = n(), .groups = "drop")

# Print the manual co-occurrence matrix
print(co_occurrence)

# Create a network graph for co-occurrences
graph <- graph_from_data_frame(co_occurrence)

# Plot the network
plot(graph, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Whale Co-occurrence Network")

```

### downsize co-occurence plot
```{r}

# Filter edges for strong co-occurrences
filtered_co_occurrence <- co_occurrence %>%
  filter(count >= 3) # Adjust threshold as needed

# Create a filtered graph
filtered_graph <- graph_from_data_frame(filtered_co_occurrence)

# Plot the filtered network
plot(filtered_graph, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Filtered Whale Co-occurrence Network")


```

### alter number of connections
```{r}

# Identify key whales based on degree (number of connections)
key_whales <- degree(graph, mode = "all") %>%
  sort(decreasing = TRUE) %>%
  head(20) # Adjust to show the top N whales

# Subgraph with only key whales
key_graph <- induced_subgraph(graph, names(key_whales))

# Plot the key whale subgraph
plot(key_graph, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Key Whale Co-occurrence Network")


```

### alter graphing type
```{r}

# Use the Fruchterman-Reingold layout
layout <- layout_with_fr(graph)

# Plot the graph 
plot(graph, layout = layout, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Improved Layout")

# Use the Kamada-Kawai layout
layout2 <- layout_with_kk(graph)

# Plot the graph 
plot(graph, layout = layout2, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Improved Layout")

# Use the Large Graph Layout 
layout3 <- layout_with_lgl(graph)

# Plot the graph with the layout
plot(graph, layout = layout3, vertex.label.cex = 0.8, edge.arrow.size = 0.5, main = "Improved Layout")

```
