---
title: Plots 01.1: Sounders' Networking
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}

# libraries
library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
## only working with the Sounders' data in this markdown file

```{r}

# figure out where you are on your computer
getwd()

# we're going to focus on only the data for the sounders for this one. Note you'll need to change the file path to match where it is on your computer
#data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/data/sounders.csv")
data <- read.csv("/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_added_rows.csv")

```

# Data Clarification
## we're going to add a column that differentiates the Sounders from the other identified gray whales
```{r}

library(dplyr)

# list of sounders
sounder_whales <- c(21, 22, 44, 49, 53, 56, 185, 356, 383, 531, 543, 723, 
                    2246, 2249, 2255, 2259, 2261, 2356, 2362, 2440, 2441)

# add a new column 'sounders' with the three possible values
data <- data %>%
  mutate(sounders = case_when(
    whale_identifier %in% sounder_whales ~ "y",           # If a sounder whale, mark as 'y'
    is.na(whale_identifier) | whale_identifier == "unknown" ~ "u", # If unknown, mark as 'u'
    TRUE ~ "n"                                           # Otherwise, mark as 'n'
  ))

# write out
#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/sounders_identified.csv", row.names = FALSE)

```

## pie chart of sounder status breakdown in this sightings grouping
### 87.5% of the sightings in this dataset are of confirmed sounders sightings (fix labels)
```{r}

# make sure your years, months, quads are being treated like characters, not numbers
data$year <- as.character(data$year)
data$month <- as.character(data$month)
data$quad <- as.character(data$quad)

# summarize the data to get the count of each 'sounders' category
sounders_summary <- data %>%
  group_by(sounders) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100,  # Calculate percentages
         label = paste0(sounders, ": ", round(percentage, 1), "%")) # Create labels

# Create the pie chart
ggplot(sounders_summary, aes(x = "", y = count, fill = sounders)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() + # Remove axes for a clean pie chart
  geom_text(aes(label = label), position = position_stack(vjust = 0.05)) +
  labs(title = "Sounder Status") +
  theme(legend.title = element_blank())

```

## counts by sounder status
### clearer picture of the breakdown - for knowledge not presentation
```{r}

ggplot(data, aes(x = sounders)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Whales by Sounder Status", x = "Sounder Status", y = "Count")


```

## seasonal variation of whales in the area
### even in the off-season they are still the most dominant - this is a nice check to ensure our data has been cleaned and structured properly.
```{r}

ggplot(data, aes(x = season, fill = sounders)) +
  geom_bar(position = "stack") +
  labs(title = "Seasonal Whale Sightings by Sounder Status", x = "Season", y = "Count")


```

## sighting trends
### this is from the subset of sightings data that have comments (2742 sightings out of over 8700 total)
```{r}

ggplot(data, aes(x = year, y = ..count.., group = sounders, color = sounders)) +
  geom_line(stat = "count") +
  labs(title = "Yearly Trends in Whale Sightings", x = "Year", y = "Count")

```

# Network Analysis (lat/ long)
## code adjusted/ updated/ fixed with ChatGPT; complete titles and explanations will follow in the next day or two
```{r}

library(dplyr)
library(igraph)
library(ggplot2)
library(ggmap)
library(tidyr)
library(purrr)

# STEP 1: Data Preparation
# Filter necessary columns
whale_subset <- data %>%
  select(whale_identifier, sounders, lat, long, sightDate) %>%
  drop_na() # Remove rows with missing data

# Identify co-occurrences: Whales spotted at the same location on the same day
co_occurrences <- whale_subset %>%
  group_by(lat, long, sightDate) %>%
  summarize(whales = list(whale_identifier), sounders = list(unique(sounders))) %>%
  ungroup()

# STEP 2: Build the Network Graph (Robust Fix)
# Filter groups with at least 2 whales and generate edges
edges <- co_occurrences %>%
  rowwise() %>%  # Ensure operations are performed row-wise
  filter(length(whales) > 1) %>%  # Keep only groups with more than one whale
  mutate(edges = list(as.data.frame(t(combn(whales, 2))))) %>%  # Generate combinations as data frame
  unnest(edges) %>%  # Expand the combinations into rows
  rename(from = V1, to = V2)  # Rename the columns for clarity

# Create graph object
network_graph <- graph_from_data_frame(edges, directed = FALSE)


# STEP 3: Analyze Spatial Trends
# Identify clusters in feeding locations
location_clusters <- whale_subset %>%
  group_by(lat, long) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# Plot feeding locations with trends over time
ggplot(location_clusters, aes(x = long, y = lat, size = count)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Frequent Feeding Locations", x = "Longitude", y = "Latitude")

# STEP 4: Visualize the Network
# Plot the network graph
plot(network_graph, vertex.size = 10, vertex.label.cex = 0.7,
     main = "Network Map of Whale Sightings")


```
## trying to simplify this using only ID'd sounders
```{r}

library(tidygraph)
library(ggraph)
library(dplyr)
library(tidyr)

# Filter for sounder whales only
sounder_data <- data %>%
  filter(sounders == "y")  # Keep only whales marked as sounders

# Group by location and date to identify co-occurrences
co_occurrences <- sounder_data %>%
  group_by(lat, long, sightDate) %>%
  summarize(whales = list(whale_identifier), .groups = "drop")  # Group whales by sighting

# Generate edges: Create pairs of whales sighted together
edges <- co_occurrences %>%
  rowwise() %>%  # Process each group individually
  filter(length(whales) > 1) %>%  # Only include groups with at least 2 whales
  mutate(edge_pairs = list(as.data.frame(t(combn(unlist(whales), 2))))) %>%  # Generate pairs
  unnest(edge_pairs) %>%  # Expand the pairs into rows
  rename(from = V1, to = V2)  # Rename columns for clarity

# Create nodes: Extract unique whales from the edges
nodes <- data.frame(
  name = unique(c(edges$from, edges$to)),  # Unique whale identifiers from edges
  stringsAsFactors = FALSE
)

# Create a tidygraph object from edges and nodes
network_graph <- tbl_graph(
  edges = edges,  # Edge list
  nodes = nodes,  # Node list
  directed = FALSE  # Undirected graph
)

# Plot the network using ggraph
ggraph(network_graph, layout = "fr") +
  geom_edge_link(aes(alpha = 0.5), show.legend = FALSE) +  # Edges
  geom_node_point(aes(color = "green"), size = 6, show.legend = FALSE) +  # Nodes
  geom_node_text(aes(label = name), repel = TRUE) +  # Labels
  theme_void() +
  labs(title = "Sounder Whale Network Graph")

```

#API Key for the maps
```{r}

#register_stadiamaps("1046a78e-35c1-44a7-8313-d79de394c168", write= TRUE)

```


## trying to enhance the graph
### ok start, not the best but in the right direction
```{r}

library(ggplot2)
library(ggraph)
library(tidygraph)
library(ggmap)
library(dplyr)

# Add latitude and longitude to the nodes data frame
nodes <- nodes %>%
  left_join(data %>% select(whale_identifier, lat, long),
            by = c("name" = "whale_identifier"))

bbox <- c(
  left = min(nodes$long, na.rm = TRUE) - 0.5,
  bottom = min(nodes$lat, na.rm = TRUE) - 1,
  right = max(nodes$long, na.rm = TRUE) + 1,
  top = max(nodes$lat, na.rm = TRUE) + 1
)

print(bbox)

# Download a base map
base_map <- get_stadiamap(
  bbox = bbox,
  zoom = 10,  # Adjust zoom as needed
  maptype = "stamen_terrain"  # Choose map type: "terrain", "toner", etc.
)

# Prepare edges for map overlay
edges <- edges %>%
  left_join(nodes %>% rename(from_lon = long, from_lat = lat), by = c("from" = "name")) %>%
  left_join(nodes %>% rename(to_lon = long, to_lat = lat), by = c("to" = "name"))

# fixing the many-to-many warning
# Check for duplicates in the nodes data frame
duplicates_in_nodes <- nodes %>%
  group_by(name) %>%
  summarize(count = n()) %>%
  filter(count > 1)

nodes <- nodes %>%
  distinct(name, .keep_all = TRUE)

# Check for duplicate edges
duplicates_in_edges <- edges %>%
  group_by(from, to) %>%
  summarize(count = n()) %>%
  filter(count > 1)

edges <- edges %>%
  distinct(from, to, .keep_all = TRUE)

edges <- edges %>%
  left_join(nodes %>% rename(from_lon = long, from_lat = lat), by = c("from" = "name")) %>%
  left_join(nodes %>% rename(to_lon = long, to_lat = lat), by = c("to" = "name"))

edges <- edges %>%
  rename(
    from_lon = from_lon.x,
    from_lat = from_lat.x,
    to_lon = to_lon.x,
    to_lat = to_lat.x
  )


# Plot the map with network overlay
first_map<- ggmap(base_map) +
  geom_segment(data = edges, aes(x = from_lon, y = from_lat, xend = to_lon, yend = to_lat),
               color = "black", alpha = 0.5) +
  geom_point(data = nodes, aes(x = long, y = lat), color = "red", size = 4) +
  geom_text(data = nodes, aes(x = long, y = lat, label = name), hjust = -0.2, size = 3) +
  theme_minimal() +
  labs(title = "Sounder Whale Network on Map",
       x = "Longitude", y = "Latitude")

print(first_map)

ggsave(filename = "/Users/cmantegna/Documents/GitHub/sicb_dp/output/sounder_network.png", plot = first_map, width = 12, height = 10, dpi = 300)

```

## chord diagram
### parameters aren't playing nice, but it does show the clustering of whales who may or may not hang out together
```{r}

library(circlize)

# Filter for sounder whales only
sounder_data <- data %>%
  filter(sounders == "y")

# Aggregate visits by whale and location
whale_location <- sounder_data %>%
  group_by(whale_identifier, lat, long) %>%
  summarize(visits = n(), .groups = "drop")

# Create a location label
whale_location <- whale_location %>%
  mutate(location = paste0("(", round(lat, 2), ", ", round(long, 2), ")"))

# Format for chord diagram
# Filter for top locations or whales by visit frequency
# Combine rare locations into "Other"
threshold <- 5  # Define a threshold for significance
edges <- edges %>%
  mutate(to = ifelse(value < threshold, "Other", to))

edges <- edges %>%
  group_by(from) %>%
  top_n(n = 10, wt = value) %>%  # Keep top 10 connections per whale
  ungroup()

# Assign colors to whales and locations
whale_colors <- setNames(colorRampPalette(c("blue", "green"))(length(unique(edges$from))), unique(edges$from))
location_colors <- setNames(colorRampPalette(c("red", "orange"))(length(unique(edges$to))), unique(edges$to))

# Combine colors
grid.col <- c(whale_colors, location_colors)

# Plot the chord diagram
chordDiagram(edges, grid.col = grid.col, annotationTrack = "grid", preAllocateTracks = 3)

# Add sector labels
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
  circos.text(
  CELL_META$xcenter, CELL_META$ylim[1] + 1.5,
  CELL_META$sector.index, facing = "clockwise", niceFacing = TRUE,
  adj = c(0, 0.5), cex = 0.7  # Decrease text size
)

}, bg.border = NA)

# Add legend for whales and locations
legend("topright", legend = c(names(whale_colors), names(location_colors)),
       fill = c(whale_colors, location_colors), title = "Legend")


```

