---
title: Junkyard
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# filter out non-identifiers
```{r}

# make the list of the exclusions
non_whale_identifiers <- c("Id", "Three", "Shrimp", "Possession Sound", "Saratoga", "Port Susan", "Stayed", "Alie Perez", "Cascadia Research", "Source", "Hat", "Hat Island", "Saw", "Presumably", "Polnell Shores", "End", "Alisa Lemire", "Brooks", "Marcie Goldberg", "Terra Parham", "We", "Gedney", "Camano Head", "Orca Network", "One", "Whidbey", "Possession Point", "North", "Mukilteo", "Most", "Little Patch", "Likely", "Is", "Individual", "Ds", "Including", "In", "Ids", "First", "North Puget", "Beh", "Spyhopping", "Looked", "Dodging", "Fluke Tip", "Identified", "Behavior", "Possible", "Island Adventures", "Blog", "Juv", "Feeding", "Fluke", "East", "Right", "Poss", "Snohomish", "No", "Fox Spit", "Sighting", "Gisborne", "Afternoon", "Everett", "Langley", "Socializing", "Near", "Sandy Point", "Alie Perex", "Delta", "Blows", "Between Hat", "Travelling", "Swimming", "Spyhop", "Spouting", "Fluke Dive", "Flips", "Newb", "Rolling", "Tail", "Deep", "Mabana", "Pebble Beach", "Within", "Also", "Elliott Bay", "Jetty Island", "Fluking", "West", "Penn Cove", "Milling")

# filter them out
timeline_data <- timeline_data %>%
  filter(!whale_identifiers %in% non_whale_identifiers)

```

# unify the identifiers
```{r}

# This is ChatGPT code to help us fix the multiple name situation

# Create a mapping of identifiers to a unified name
identifier_mapping <- c(
  "CRC-21" = "Shackleton",
  "#21" = "Shackleton",
  "Shackleton" = "Shackleton"
  # Add more mappings as needed
)

# Replace identifiers in the data using the mapping
timeline_data$whale_identifiers <- recode(timeline_data$whale_identifiers, !!!identifier_mapping)

```

# extract identifiers
```{r}

# Function to extract whale identifiers
extract_whale_identifiers <- function(comment) {
  pattern <- "(#[\\d]+|CRC-\\d+|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"  # Matches #723, CRC-44, or names like Shakleton
  str_extract_all(comment, pattern)
}
```

# more identifiers stuff before I manually cleaned the data; these steps are needed for the bulk cleaning, but don't work with the fine-tuning 

### this step is for cleaning data from the beginning - use lines X-X to dis-aggregate the identifiers from the file named "\~/data_adjusted/sounders_clean_noIdentifiers.csv".

```{r}

# we're going to extract all of the comments from the column and make them individual rows so we can assess what we have. This code came directly from ChatGPT

library(dplyr)
library(tidyr)
library(stringr)

# Function to extract whale identifiers, including plain numbers and case-insensitive CRC identifiers
# Updated pattern:
  # 1. Matches # followed by digits (#123)
  # 2. Matches case-insensitive CRC followed by digits (CRC-123, crc123)
  # 3. Matches plain numbers (1234)
  # 4. Matches names with one or two capitalized words (e.g., Shackleton, Little Patch)
extract_whale_identifiers <- function(comment) {
  pattern <- "(#[\\d]+|(?:[Cc][Rr][Cc][-]?\\d+)|\\b\\d+\\b|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"
  str_extract_all(comment, pattern)
}

# Apply the extraction function to create a new column
data <- data %>%
  mutate(whale_identifiers = sapply(lapply(comments, extract_whale_identifiers), paste, collapse = ", "))


# Filter for rows where whale_identifiers is not NA
#timeline_data <- data %>% filter(!is.na(whale_identifiers))

# check your work & write out as a table to save progress if you need to.
head(data)

#write.csv(data, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/timeline_data.csv", row.names = FALSE)

```

## list identified whales, unique names

```{r}

# we have lists of IDs in the comments column, but not all of them are actual IDs but rather observation note (i.e., young whale) and we want to know what we're working with. This code came from ChatGPT.

library(stringr)

# Define the regex pattern used for extraction
pattern <- "(#[\\d]+|CRC-\\d+|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"

# Extract identifiers from comments
extracted_identifiers <- str_extract_all(data$comments, pattern)

# Flatten the list and get unique values
unique_identifiers <- unique(unlist(extracted_identifiers))

# View the unique identifiers
print(unique_identifiers)

# write out to your data_adjusted output folder

# unblock this if you have made changes before this step to overwrite the old list
#write.csv(unique_identifiers, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/identifiers.csv", row.names = FALSE)

```

## unique identifiers and sighting separations with counts in-tact. **This is the next step after line 93.**

## Data Cleaning, Part 2

## remove non-whale identifiers

```{r}

# Define a list of valid whale identifiers
valid_identifiers <- c("#723", "#49", "#383", "CRC-44", "CRC-56", "#44", "Shrimp", "Gray", "#185", "Shackleton", "#21", "Earhart", "#22","CRC-49", "CRC-53", "Saratoga Gray", "#724", "CRC-531", "CRC-383", "CRC-21", "Possible Gray", "#531", "CRC-185", "Patch", "Lucyfer", "Dubknuck", "Lucyfe", "#56", "Gray Whale", "Little Patch", "#53", "Shack", "Newbie", "Likely Shackleton", "Sound Grey", "Whale", "#2259", "Shackelton", "Dubnunk", "#2258", "#356", "Possibly Dubknuck", "CRC-2246", "Dubnuk", "Dubnuck", "Juv", "CRC-2262", "Lucy", "#2246", "Gray Whales", "Earheart", "Stardust", "Hattie", "Sounder", "Thidwick", "Gretchen", "Cascade", "Carson", "Tahoma", "Stalwart", "Eaerhart", "Gisborne", "CRC-723", "CRC-22", "CRC-2549", "Delta", "CRC-2234", "Crc", "Cretchen", "CRC-2255", "#2262", "Newb", "Adult Grays", "Lucifer", "#2249", "Patches", "#42", "#33", "#2356", "#2261", "#2253", "#2252", "#2251", "#2234", "PCFG 185", "CRC2246", "young gray whale", "patch", "little patch", "53", "49", "56", "352", "185", "723")

# Filter whale_identifiers to keep only valid ones
data <- data %>%
  mutate(whale_identifiers = lapply(whale_identifiers, function(ids) {
    ids[ids %in% valid_identifiers]
  }))

```

## re-check for unique identifiers

## list identified whales, unique names

```{r}

# we have lists of IDs in the comments column, but not all of them are actual IDs but rather observation note (i.e., young whale) and we want to know what we're working with. This code came from ChatGPT.

library(stringr)

# Define the regex pattern used for extraction
pattern <- "(#[\\d]+|CRC-\\d+|[A-Z][a-z]+(?:\\s[A-Z][a-z]+)?)"

# Extract identifiers from comments
extracted_identifiers <- str_extract_all(data$comments, pattern)

# Flatten the list and get unique values
unique_identifiers <- unique(unlist(extracted_identifiers))

# View the unique identifiers
print(unique_identifiers)

# write out to your data_adjusted output folder

# unblock this if you have made changes before this step to overwrite the old list
#write.csv(unique_identifiers, "/Users/cmantegna/Documents/GitHub/sicb_dp/output/data_adjusted/identifiers2.csv", row.names = FALSE)

```

## unify multiple identifiers for an individual

```{r}

# Create a synonym mapping (key-value pairs)
identifier_mapping <- c("A1" = "A", "A2" = "A", "B1" = "B", "B2" = "B")  # Example

# Standardize identifiers
data <- data %>%
  mutate(whale_identifiers = lapply(whale_identifiers, function(ids) {
    sapply(ids, function(id) ifelse(id %in% names(identifier_mapping), identifier_mapping[id], id))
  }))

```

## adjust the df to account for multiple whales listed in one comment line

```{r}

library(tidyr)

# Assume the `count` column represents the total number of whales observed
data <- data %>%
  unnest(whale_identifiers) %>%  # Each identifier gets its own row
  group_by(row_id = row_number()) %>%  # Add unique row identifier
  mutate(count = count / n()) %>%  # Divide count by number of identifiers
  ungroup() %>%
  select(-row_id)  # Remove helper column

```

# viz
## a mapping thing
```{r}

# plotting with lat/ long coordinates, but not overlayed on a map - just want to see what it looks like

library(ggplot2)

ggplot(data, aes(x = long, y = lat, color = whale_identifiers)) +
  geom_point(alpha = 0.6) +
  labs(title = "Map of Whale Sightings", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```

## timeline of sightings

```{r}

# convert sightDate to Date format (if it wasn't done earlier)
#timeline_data$sightDate <- as.Date(timeline_data$sightDate, format = "%m/%d/%y")

ggplot(timeline_data, aes(x = sightDate, y = whale_identifiers)) +
  geom_point(aes(color = whale_identifiers), alpha = 0.6) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Timeline of Whale Sightings", x = "Date", y = "Whale Identifiers", color = "Whale Identifiers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## movement patterns

```{r}

# Sort by identifiers and date
timeline_data <- timeline_data[order(timeline_data$whale_identifiers, timeline_data$sightDate), ]

ggplot(timeline_data, aes(x = long, y = lat, group = whale_identifiers, color = whale_identifiers)) +
  geom_path(alpha = 0.6) +
  geom_point(alpha = 0.6) +
  labs(title = "Movement Patterns of Identifiable Whales", x = "Longitude", y = "Latitude", color = "Whale Identifiers") +
  theme_minimal() +
  theme(legend.position = "right")

```
